---
title: "Example 2: Delayed treatment effect"
author: "Godwin Yung"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 2: Delayed treatment effect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

We provide here code to replicate Figures 3 and 4A from Example 2 of our manuscript. We rely on R packages tidyverse, plyr, and ggplot2 for cleaner coding and advanced visualizations.

```{r, message=FALSE, warning=FALSE}
library(npsurvSS)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
```

## Figure 3
Suppose 240 active and 120 control patients are enrolled over the course of 14 months and followed for 11 months. Assume that accrual follows a truncated exponential distribution with shape parameter 0.1; loss to follow-up follows a Weibull distribution with shape parameter 2 and scale parameter $log(1/0.99)^{1/2}/25$, so that 1% of patients are permanently censored over 25 months of follow-up; and survivals follow exponential distributions with 9 and 6 month medians. All these information are captured in the following "arm" objects:
```{r}
# Active
arm1 <- create_arm(size=240,
                   accr_time=14,
                   accr_dist="truncexp",
                   accr_param=0.1,
                   surv_scale=per2haz(9),
                   loss_scale=log(1/0.99)^(1/2)/25,
                   loss_shape=2,
                   total_time=25)

# Control
arm0 <- create_arm(size=120,
                   accr_time=14,
                   accr_dist="truncexp",
                   accr_param=0.1,
                   surv_scale=per2haz(6),
                   loss_scale=log(1/0.99)^(1/2)/25,
                   loss_shape=2,
                   total_time=25)
```
To visualize accrual and loss-to-follow-up:
```{r, fig.show='hold'}
# Accrual
tibble(
  x = seq(0, 14, 0.1),
  y = paccr(x, arm0)
) %>%
  ggplot(aes(x, y)) +
  geom_line() +
  labs(x = "Time from first patient in (months)",
       y = "Accrual CDF")

# Loss to follow-up
tibble(
  x = seq(0, 25, 0.1),
  y = ploss(x, arm0)
) %>%
  ggplot(aes(x, y)) +
  geom_line() +
  labs(x = "Time from study entry (months)",
       y = "Loss to follow-up CDF")
```

In the manuscript, we explore the impact on power when survival in the active arm follows instead a 2-piece exponential distribution, where the first piece overlaps with the control arm and the second piece deviates in such a way that median survival remains at 9 months. Denoting the changepoint by $\tau_1$ and the arm-specific hazard rates by $\lambda_0$ and $(\lambda_{11}, \lambda_{12})=(\lambda_0, \lambda_{12})$, simple algebra dictates that $\lambda_{12}=\{\lambda: e^{-\lambda(9-\tau_1)}=\frac{0.5}{e^{-\lambda_0 \tau_1}}\}$. We can utilize functions in npsurvSS to calculate and visualize survival in the active arm under various changepoints:
```{r, fig.width=5}
# Calculate survival curves
x.vec     <- seq(0, 25, 0.1)  # vector of unique x-coordinates
tau1.vec  <- seq(0, 4.5, 1.5) # vector of unique changepoints
arm1t     <- arm1             # initialize active arm
y         <- c()              # vector or all y-coordinates
for (tau1 in tau1.vec) {
  # Update scale and interval parameters for the active arm
  arm1t$surv_scale    <- c(arm0$surv_scale[1], per2haz(9-tau1, 1-0.5/psurv(tau1, arm0, lower.tail=F)))
  arm1t$surv_interval <- c(0, tau1, Inf)
  
  # Calculate y-coordinates
  y <- c(y, psurv(x.vec, arm1t, lower.tail=F))
}

# Visualize survival curves
tibble(
  tau1 = rep(tau1.vec, each=length(x.vec)), 
  x = rep(x.vec, length(tau1.vec)),
  y = y
) %>%
  ggplot(aes(x, y)) +
  geom_line(aes(color=factor(tau1), lty=factor(tau1))) +
  labs(x = "Time from study entry (months)",
       y = "Survival function",
       color = "Changepoint",
       lty = "Changepoint")
```

## Figure 4A
Finally, to calculate power for various non-parametric tests, we take advantage of `power_two_arm`'s ability to accomodate multiple tests at a time. For the sake of efficiency here, we will only consider changepoints 0, 0.5, 1, ..., 5.5. In the manuscript, we consider changepoints 0, 0.1, 0.2, ..., 5.9.
```{r}
tau1.vec <- seq(0, 5.5, 0.5) # vector of changepoints
table_4a <- data.frame(matrix(0, nrow=length(tau1.vec), ncol=7)) # initialize results table
for (r in 1:length(tau1.vec)) {
  
  tau1 <- tau1.vec[r]
  
  # Update scale and interval parameters for the active arm
  arm1t$surv_scale    <- c(arm0$surv_scale[1], per2haz(9-tau1, 1-0.5/psurv(tau1, arm0, lower.tail=F)))
  arm1t$surv_interval <- c(0, tau1, Inf)
  
  # Calculate power and store results
  table_4a[r,] <- c(tau1,
                    power_two_arm(arm0, arm1t,
                                  test = list(list(test="weighted logrank"),
                                              list(test="weighted logrank", weight="n"),
                                              list(test="weighted logrank", weight="FH_p1_q1"),
                                              list(test="survival difference", milestone=11),
                                              list(test="rmst difference", milestone=11),
                                              list(test="percentile difference", percentile=0.5)))$power)
  
}

# Convert table to long format and re-label tests
table_4a <- gather(table_4a, "test", "power", 2:7) %>%
  mutate(test = recode(test, X2 = "wlogrank (1)",
                       X3 = "wlogrank (GB)",
                       X4 = "wlogrank (FH)",
                       X5 = "11-month surv diff",
                       X6 = "11-month rmst diff",
                       X7 = "median diff")) %>%
  as_tibble()
names(table_4a)[1] <- "tau1"
```
The resulting table looks like this:
```{r}
table_4a
```
It can be used to generate Figure 4A in the manuscript:
```{r, fig.width=5}
ggplot(table_4a, aes(x=tau1, y=power)) +
  geom_line(aes(color=test, lty=test)) +
  labs(x = "tau1",
       y = "Power",
       color = "Test",
       lty = "Test")
```
